---
title: "Occupancy detection in RMV2.0 TOWT model"
output: html_notebook
---

In this notebook, we try to visualize the automatic occupancy detection
algorithm used in the RMV2.0 time-of-week temperature (TOWT) model. This
article does not promote use of the algorithm for all cases. You should
determine whether it is appropriate for your building and usage history.

# Intro

The occupancy detection is done by `findOccUnocc()`. Comments in the code
explain the proccess:

<blockquote>
Define 'occupied' and 'unoccupied' based on a regression
of load on outdoor temperature: times of week that the regression usually
underpredicts the load will be called 'occupied', the rest are 'unoccupied'
This is not foolproof but usually works well.
</blockquote>
<blockquote>
If the regression underpredicts the load more than 65% of the time
then assume it's an occupied period.
</blockquote>

Some details are important to clarify:

* These regression fit used here is thrown away after occupancy detection. It
  is not related to the baseline model we eventually get.
* We use a two-changepoint model to fit the 8,760-hour data against temperature,
  and no additional variables. 
* _Underpredict_ means residual > 0, where residual = data - regression prediction.
* _65% of the time_: At this point, we group data points by time of week, so
  there are 7*24 = 168 groups. For each group, we then calculate the number of
  data points (1 per week, usually 52 total) and of those, the number of data
  points where the regression underpredicts. If the ratio is greater than 65%
  (an arbitrary threshold), we set this time of week as "occupied".

Here is the call stack:

* `train_model` (called at <tt>server.R#298</tt>, defined at <tt>utils.R#215</tt>)
* `towt_baseline` (called at <tt>utils.R#243</tt>, defined at
  <tt>towt_baseline.R#42</tt>)
* `makeBaseline` (called at <tt>towt_baseline.R#74</tt>, defined at
  <tt>towt_baseline.R#433</tt>)
* `fitLBNLregress` (called at <tt>towt_baseline.R#491</tt>, defined at
  <tt>towt_baseline.R#224</tt>)
* `findOccUnocc` (called at <tt>towt_baseline.R#296</tt>, defined at
  <tt>towt_baseline.R#186</tt>)

# Some development work

Can we recover enough data to show the process, from a saved project file? If
not, at what point in the code do we need to store some more data?

```{r}
library(dplyr)
library(glue)
library(RMV2.0)
```

```{r}
# The 'project' file saved when you used the RMV2.0 add-in.
rds_file <- "C:/RMV2.0 Workshop/something/Project_02.19.rds"
Project <- readRDS(rds_file)
i = 1
```

What next? Here, we recreate the function call to `towt_baseline`.

```{r}

pre_Data_i <- Project$Data_pre[[i]]
timescaleDays <- Project$model_obj_list$models_list[[i]]$towt_model$timescaleDays
intervalMinutes <- Project$Data_pre_summary[1,6]
fahrenheit <- Project$fahrenheit

res_baseline <- towt_baseline(train_Data = pre_Data_i,
                              pred_Data = pre_Data_i,
                              timescaleDays = timescaleDays,
                              intervalMinutes = intervalMinutes,
                              fahrenheit = fahrenheit,
                              )
```

Now, to illustrate some of the data structures used in the next step ...

```{r}
train <- Project$model_obj_list$models_list[[i]]$train
train
```

```{r}
train$time <- as.POSIXlt(train$time, format="%m/%d/%y %H:%M")
head(train$time)
```

```{r}
pred <- Project$model_obj_list$models_list[[i]]$pred
pred
```

```{r}
pred$time <- as.POSIXlt(pred$time, format="%m/%d/%y %H:%M")
head(pred$time)
```

What next? Here, we recreate the function call to `makeBaseline`. The function loops
over a list of timestamps spaced by timescaleDays (the hyperparameter set by the user
in the GUI, from 15 to 90 days). At each step, it sets up weights centered at the
timestamp and calls `fitLBNLregress` to create a mini-regression. It then bundles all
these regressions together.

```{r}
verbosity=5
towt_model <- makeBaseline(train$time,
                             train$eload,
                             train$Temp,
                             pred$time,
                             pred$Temp,
                             intervalMinutes=intervalMinutes,
                             timescaleDays=timescaleDays,
                             fahrenheit=fahrenheit,
                             verbose=verbosity)

```

What next? Here, we recreate the function call to `fitLBNLregress`.

```{r}
dataTime <- train$time
dataLoad <- train$eload
dataTemp <- train$Temp
predTime <- pred$time
predTemp <- pred$Temp

tempKnots = (c(40, 55, 65, 80, 90)-32)*5/9
doTemperatureModel<-T
verbose<-5

npoints = length(dataLoad)
t0 = min(dataTime,na.rm=T)
t1 = max(dataTime,na.rm=T)
deltaT = as.numeric(difftime(t1,t0,units="days"))
nsegments = max(1,ceiling(deltaT/timescaleDays))
segmentwidth = (npoints-1)/nsegments
pointlist = floor(sort(npoints-segmentwidth*(0:nsegments))+0.001)
nModelRuns = max(1,length(pointlist))

#for (irun in 1:nModelRuns)
irun <- 1
tcenter = dataTime[pointlist[irun]]
tDiff = as.numeric(difftime(tcenter,dataTime,units="days"))
tDiffPred = as.numeric(difftime(tcenter,predTime,units="days"))
weightvec = timescaleDays^2/(timescaleDays^2 + tDiff^2)

regOut = fitLBNLregress(dataTime, dataLoad, dataTemp, predTime, predTemp,
			tempKnots = tempKnots, weightvec=weightvec,
			intervalMinutes=intervalMinutes,fahrenheit=fahrenheit,
			doTemperatureModel=doTemperatureModel,verbose=verbose)
```

What next? Here, we recreate the function call to `findOccUnocc` (from the first step
inside `fitLBNLregress`). Note:

* This is where we first label each timestamp with an integer `intervalOfWeek` using the weekday, hour, and minute parts of the timestamp data.
* The `intervalOfWeek` increments by 1 every `intervalMinutes` (in this project, every 60 minutes).
* Eg., Sunday, 2006-01-01T00:00 has (wday, hour, minute) = (0,0,0). 00:00 through 00:59 would all be labeled `intervalOfWeek` = 1.
* Eg., Sunday, 2006-01-01T01:00 has (wday, hour, minute) = (0,1,0), which is mapped to intervalOfWeek = 2.

```{r}
timeVec <- dataTime
loadVec <- dataLoad
tempVec <- dataTemp
#predTime
#predTemp
#tempKnots
#weightvec
#intervalMinutes
#fahrenheit
#doTemperatureModel
#verbose

minuteOfWeek = 24*60*timeVec$wday+60*timeVec$hour + timeVec$min
intervalOfWeek = 1+floor(minuteOfWeek/intervalMinutes)

# If we have temperature data then fit the time-of-week-and-temperature model

if (fahrenheit) {
	# temperature vector is already in fahrenheit
	tempVecF = tempVec
	tempVec = (tempVec-32)*5/9
	tempVecPredF = predTemp
	tempVecPred = (predTemp-32)*5/9
} else {
	tempVecF = (tempVec*9/5)+32
	tempVecPredF = (predTemp*9/5)+32
	tempVecPred = predTemp
}

# findOccUnocc requires Fahrenheit temperatures; everywhere else we can use either
#  Celsius or Fahrenheit, as long as temperature knots are set appropriately
#
# base occupied/unoccupied decision only on cases where we have load data:
okload = !is.na(loadVec)
occInfo = findOccUnocc(intervalOfWeek[okload],loadVec[okload],tempVecF[okload])
head(occInfo,40)
tail(occInfo,20)
```

What next? Here, we demonstrate the occupancy detection algorithm within `findOccUnocc`. Note:

* `uTOW` = unique time-of-week. If `intervalMinutes`=60, then there are up to 168 such
numbers. We do not need to sort them, which is unnecessary.
* 

```{r}
intervalOfWeek2 <- intervalOfWeek[okload]
loadVec2 <- loadVec[okload]
TempF <- tempVecF[okload]
#intervalMinutes
#verbose

# Figure out which times of week a building is in one of two modes
#  (called 'occupied' or 'unoccupied')

# RMV2.0 does not sort this vector. Although that doesn't matter,
# I prefer to have it sorted now.
#uTOW = unique(intervalOfWeek2)
uTOW = sort(unique(intervalOfWeek2))
nTOW = length(uTOW)

# Define 'occupied' and 'unoccupied' based on a regression
# of load on outdoor temperature: times of week that the regression usually
# underpredicts the load will be called 'occupied', the rest are 'unoccupied'
# This is not foolproof but usually works well.
#
TempF50 = TempF-50
TempF50[TempF > 50] = 0
TempF65 = TempF-65
TempF65[TempF < 65] = 0

if (verbose > 4) {
	cat("Fitting temperature regression...\n")
}
amod = lm(loadVec2 ~ TempF50+TempF65,na.action=na.exclude)

okocc = rep(0,nTOW)
cat("Detecting occupancy ...\n")
for (itow in 1:nTOW) {
	okTOW = intervalOfWeek2==uTOW[itow]
	# if the regression underpredicts the load more than 65% of the time
	# then assume it's an occupied period
	if ( sum(residuals(amod)[okTOW]>0,na.rm=T) > 0.65*sum(okTOW) ) {
		okocc[itow]=1
	}
  cat(glue('[{format(t,width=3)}] {format(nunder,width=2)}/{format(ntotal,width=2)} -> {occ}',
           t=uTOW[itow],
           nunder=sum(residuals(amod)[okTOW]>0,na.rm=T),
           ntotal=sum(okTOW),
           occ=okocc[itow]
           ),'\n')

}
occInfo = cbind(uTOW,okocc)
```

# Visualization
Here are a bunch of ways to visualize the occupancy detection algorithm. The best one
is all the way at the bottom, which is mostly scripted in Javascript with Charts.js.

## Static plot using ggplot2

```{r}
library(ggplot2)
tow <- 40
okTOW = intervalOfWeek2==tow

ggplot() +
  geom_point(data=data.frame(x=TempF,y=loadVec2), aes(x, y), color='gray') +
  geom_point(data=data.frame(x=TempF[okTOW],y=loadVec2[okTOW]), aes(x, y), color='blue') +
  geom_line(data=data.frame(x=TempF,y=predict(amod)), aes(x, y)) +
  xlab("Temperature (°F)") + #"\xB0"
  ylab("Load (kW)")

```


## A bunch of static plots

```{r}
# To repeat the same data in every panel, simply construct a data frame
# that does not contain the faceting variable.
mydata=data.frame(x=TempF,y=loadVec2,ypred=predict(amod),tow=intervalOfWeek2,timeVec=timeVec)
mydata[mydata$tow<10,]
```

```{r}
ggplot(mydata[mydata$tow<10,], aes(x, y)) +
  geom_point(data = transform(mydata, tow = NULL), colour = "grey85") +
  geom_point() +
  facet_wrap(vars(tow))
```

## Animated plot using plotly

```{r}
library(plotly)

mydata %>%
  plot_ly(
    x = ~x,
    y = ~y,
    frame = ~tow,
    type = 'scatter',
    mode = 'markers',
    showlegend = F
  )
```
## Using plotly to animate ggplot2 style

```{r}
#  geom_point(aes(x, y), color='gray') +
#  geom_line(aes(x, ypred)) +

p <- ggplot(mydata) +
  geom_point(aes(x, y, frame=tow), color='blue') +
  xlab("Temperature (°F)") + #"\xB0"
  ylab("Load (kW)")

fig <- ggplotly(p) %>%
  animation_opts(
    frame=500, easing='linear', redraw = F
  )
fig
```

## Plotly animation WITH full point cloud in the background
This suffers from slow speed (heavy on the CPU usage).

```{r}
mydata %>%
  plot_ly(x=~x,mode='none') %>%
  layout(xaxis=list(title=list(text='Temperature (°F)')),
         yaxis=list(title=list(text='Load (kW)'))) %>%
  add_trace(y=~y,name='loads', type='scatter', mode='markers',
            showlegend=F, marker=list(color='grey'), text = 'none') %>%
  add_lines(y=~ypred,name='fit',
            showlegend=F, line=list(color = 'black')) %>%
  add_trace(
    y = ~y,
    frame = ~tow,
    type = 'scatter',
    mode = 'markers',
    showlegend = F,
    marker = list(color = 'blue')
  ) %>%
  animation_opts(
    frame=500, easing='linear', redraw=F
  )
```

## Customized widget with Chart.js

Next, I've created a javascript-based charting widget because this is more
interactive than an animation, and works better because it doesn't
animate the point cloud in the background. The widget does not yet interact with R,
so I'm just going to show how to export some data to a file that I have manually pasted
into the widget. Here are some future features:

* Generate the widget from the R markdown notebook automatically.
* Add a slider for the detection threshold (now at 65%).
* Caution: the time-of-week numbers here may not be the same as in the R script.

Here is how I export the data to the widget.

```{r}
# Convert to JSON for Chart.js
dataByTOW=uTOW %>% purrr::map(~ mydata[mydata$tow==.x,])

#fitLine = unique(mydata[with(mydata,order(x)),c('x','ypred')])
fitLine = data.frame(x=TempF,y=predict(amod)) %>% unique() %>% arrange(x)

# Here's how to save the data directly as a block of javascript code.
# This is a temporary hack to pass the data to a widget that is just hand-coded HTML.
dataByTOWjson = jsonlite::toJSON(dataByTOW, pretty=F)
fitLinejson = jsonlite::toJSON(fitLine, pretty=F)
sink('datafile.js')
cat(glue('const dataByTOW = {dataByTOWjson};
const fitLine = {fitLinejson};'))
sink()
#mydata %>% mutate(x=x,y=ypred)

# Here's how to store the data in native R objects.
# This is for testing the next version of the widget, using htmlwidgets package.
saveRDS(list(dataByTOW=dataByTOW, fitLine=fitLine, mydata=mydata), file="mydatafile.RDS")
```

Here is the widget. Try pointing the mouse over the time-of-week grid:

* Vertical: from Sunday through Saturday
* Horizontal: from midnight to midnight
* Collapsed on: week of the year

Here is the old code to insert an iframe. It should not actually render one.

```
iframe width="900" height="1200" src="./mychart3.html" /iframe
```

```{r}
library(mywidget)
#occupancy_widget("Hello, globals!", dataByTOWjson, fitLinejson, 'json')
occupancy_widget("hello, solar system!",dataByTOW, fitLine, 'r', width='auto', height='auto')
```


That is all. Thank you for reading.
Find more from [my fork of RMV2.0](https://github.com/simularis/RMV2.0/tree/dev/demo)
